Useful R tidbits: Social Networks

### 2022-02-16 _academics/ps753-glossary/ps753-glossary.Rmd
knits

(closeness) centrality
: (forthcoming)

(betweenness) centrality
: (forthcoming)

(eigenvector) centrality
: (forthcoming)

@@

>>>>

@@

## Week 4 - Status & Eigenvectors

### Class lecture 4a: [Status and Hierarchy](https://www.loom.com/share/91c521e65705482cb7249140be3b0274)

closeness centrality
: the sum of geodesic distances (shortest paths) to all other points in the graph. Divide by $n-1$, then invert. A measure of how close a node is to all of the other nodes in the network.

### Class lecture 4b: [Status and Prestige](https://www.loom.com/share/e0484b4af1bb4d3b8242715c8cba30f8)

prestige
: signal of the quality of a node, of a node's visibility within the network

eigenvector centrality
: "takes into account the centrality of other nodes to which a node is connected. That is, being connected to other central nodes increases centrality." Takes into account the centrality of the nodes a node is connected to, considering the importance of the connected nodes and not just their quantity or path length.
$$\lambda \text{c} = \text{Wc}$$

Bonacich Power
: "A closely related concept, but includes a weighting factor that emphasizes global vs. local structure (and negative connections)"
$$\text{c}(\alpha, \beta) = \alpha(\text{I} - \beta\text{W})^{-1}\text{W1}$$

### Class lecture 4c: [Hubs and Bridges](https://www.loom.com/share/e85fa1918f70438e9bb67232ade95594)

bridge
: a node with few ties to central actors

hub
: a node with many ties to peripheral actors

reflected centrality
: the degree of centrality that a node receives from a connected node that is due to that node's centrality. If A has influential friend B, then A's eigenvector centrality will be higher; reflected centrality is the portion of A's centrality that comes from B's centrality.

derived centrality
: the remainder of the eigenvector centrality that A receives from B in the example above, that is not due to B's centrality but just from B being a connected node

Reflected and derived centrality can be represented in the following matrix by Mizruchi et al:

||High reflected centrality|Low reflected centrality|
|---|---|---|
|High derived centrality|Cosmopolitans|Pure bridges|
|Low derived centrality|Pure hubs|Peripherals|

The four prototypes are:

cosmopolitans
: well-connected nodes that are connected to other well-connected nodes; the "cool kids" table

pure hubs
: well-connected nodes that are connected to nodes that are not well-connected; the cool kid who talks to uncool kids

pure bridge
: a node connected to few but high-centrality nodes; a friend of the cool kid with few other friends

peripheral
: a node with few and low-centrality connections; me and my D&D nerd friends in high school

### Class lecture 4d: [Status and Core/Periphery Networks](https://www.loom.com/share/2b5fdf986b164bb785a4c8c0fca0d1e9)

betweenness centrality
: (forthcoming)

### 2022-02-16 _academics/ps753-useful-r-tidbits/ps753-useful-r-tidbits.Rmd
knits

# retrieve the index of isolate nodes
# (nodes with component count of 1 from "components" above)
isolates(gotbook.stat)

@@

Could also call it with other `sna` functions like `betweenness`, `closeness`; tutorial references `eigenvalue` but there is no such function in `sna`. `evcent` perhaps?

@@

>>>>

# retrieve the index of isolate nodes
# (nodes with component count of 1 from "components" above)
isolates(gotbook.stat)
# There is no direct command in igraph, but we can do this:
# Create a list of the degree of each node in the network
deg_counts <- igraph::degree(gotbook.ig, loops = FALSE)
# filter and count the nodes with 0 degrees (or any other quantity of interest)
length(deg_counts[deg_counts == 0])

@@

Could also call it with other `sna` functions like `betweenness`, `closeness`

@@

### Eigenvector centralization

`statnet` uses `evcent()` to calculate the eigenvalue centrality score for each node in the network:

```r
evcent(imf2014.stat, ignore.eval=TRUE))
```

Eigenvector centrality index for the network:

```r
centralization(imf2014.stat, evcent)
```

In `igraph`, a set of eigenvector-related information is created with `centr_eigen()`:

```r
# If the network is directed, specify "directed - T" - will not auto-detect
eigen_info <- centr_eigen(imf2014.ig, directed = T)
# Centrality score for node 3:
eigen_info[3]$vector
# Eigenveector centrality index for the network:
eigen_info$centralization
```

**The scores calculated by `igraph` and `statnet` are different. We aren't sure why.** It appears that igraph [counts incoming ties to calculate eigenvector centrality](https://stackoverflow.com/questions/21035598/what-is-evcent-returning-for-directed-graphs), and statnet [recommends using Bonachic power instead for directed networks](http://mailman13.u.washington.edu/pipermail/statnet_help/2015/002170.html).

### Bonacich Power Centrality

`igraph`:

```r
power_centrality(imf2014.ig)
```

`statnet`:

```r
bonpow(imf2014.stat)
```

Again, there appear to be some inconsistency between `igraph` and `statnet` in the calculations, with statnet apparently not incorporating weights and failing on singluar matrices.

### Derived and Reflected Centrality

There are no library routines for these calculations. Convert the data to a matrix first:

```r
mat2014 <- as.matrix(as_adjacency_matrix(imf2014.ig, attr="weight"))
```

> To calculate the proportion of centrality that is received, we first square the adjacency matrix. The diagonal of the adjacency matrix is equal to the the square of node degree. We then divide this diagonal (sqared degree) by total sqaured indegree (calculated by rowSums) to get the proportion of received centrality.

```r
# square the adjacency matrix
mat2014sq <- t(mat2014) %*% mat2014

# Calculate the proportion of reflected centrality.
imf2014.nodes$rc <- diag(mat2014sq) / rowSums(mat2014sq)

# freplace missing values with 0
imf2014.nodes$rc <- ifelse(is.nan(imf2014.nodes$rc), 0, imf2014.nodes$rc)

# Calculate received eigenvalue centrality
imf2014.nodes$eigen.rc <- imf2014.nodes$eigen * imf2014.nodes$rc
```

> If total centraltiy is 1, then derived centrality is simply 1 - the proportion of eigenvector centrality due to received centrality.

```r
# Calculate the proportion of derived centrality.
imf2014.nodes$dc <- 1 - diag(mat2014sq) / rowSums(mat2014sq)

# replace missing values with 0
imf2014.nodes$dc <- ifelse(is.nan(imf2014.nodes$dc), 1, imf2014.nodes$dc)

# Calculate received eigenvalue centrality
imf2014.nodes$eigen.dc <- imf2014.nodes$eigen * imf2014.nodes$dc
```
